---
title: "Lab 6"
author: "Dominic LaRoche"
date: "Tuesday, October 14, 2014"
output: word_document
---

```{r readin}
rm(list=ls())
library(sp)
library(gstat)  #load the library
library(lattice)
library(pander)
data(meuse.grid)
data(meuse)
#str(meuse.grid)

hist(meuse$lead, breaks=12)
hist(log10(meuse$lead), breaks=12)
hist(meuse$zinc, breaks=12)
hist(log10(meuse$zinc), breaks=12)
hist(meuse$om, breaks=12)
hist(sqrt(meuse$om), breaks=12)
#ln tranformation of om won't really help here but sqrt seems to help stabilize the variance
#Extract a subset of the target variable
meuse.pb<-meuse[seq(1, length(meuse$lead), by=3), c('x', 'y', 'lead', 'zinc', 'om')]
#str(meuse.pb)
# rownames(meuse.pb) 
 
meuse.pb<-cbind(meuse.pb, ltpb=log10(meuse.pb$lead), ltzn=log10(meuse.pb$zinc))
#Create a data frame of the lead observations at the extra points that have not been included in the subsample
meuse.extra <- meuse[setdiff(rownames(meuse), rownames(meuse.pb)), c("x", "y", "lead")]
meuse.extra <- cbind(meuse.extra, ltpb = log10(meuse.extra$lead))
coordinates(meuse)<- ~x+y  #convert data frame to "SpatialPointsDataFrame"
coordinates(meuse.pb) <- ~ x + y
coordinates(meuse.extra) <- ~ x + y
coordinates(meuse.grid) <- ~ x + y
# class(meuse)
# summary(meuse.pb)
bubble (meuse.pb, zcol= 'ltpb', main= 'log10 lead distribution')
```

```{r OrdinaryKriging}
#variogram construction
v.ltpb<-variogram(ltpb~1, data=meuse.pb, cutoff=1800, width=200)
plot(v.ltpb)
m.ltpb<-vgm(psill=0.08, model='Sph', range =800, nugget=0.03)
plot(v.ltpb, pl=T, model=m.ltpb)
m.ltpb.f <- fit.variogram(v.ltpb, m.ltpb)
plot(v.ltpb, pl=T, model=m.ltpb.f)

#conduct ordinary kriging
k.o <- krige(ltpb ~1, locations=meuse.pb, newdata=meuse.grid, model=m.ltpb.f)
# summary statistics
summary(k.o)

#validation: prediction at the extra points
k <- krige(ltpb ~ 1, meuse.pb, meuse.extra, m.ltpb.f)
# compute and summarize validation errors
summary(k)
diff <- k$var1.pred - meuse.extra$ltpb
summary(diff)
MSE_ok<-sqrt(sum(diff^2)/length(diff))
Bias_ok<-sum(diff)/length(diff)     # mean error (bias)
med_ok<-median(diff)     # median error
```


```{r CrossValKriging,message=FALSE,results='hide'}
cv_ok<-krige.cv(ltpb ~1, locations=meuse.pb,  model=m.ltpb.f, nfold=52)
MSE_okcv<-sqrt(sum(cv_ok$residual^2)/length(cv_ok$residual))
Bias_okcv<-sum(cv_ok$residual)/length(cv_ok$residual)
med_okcv<-median(cv_ok$residual)
#Created my own Loop for n-fold cross validation of ordinary kriging
#I wanted to see whether this would produce similar results 
data(meuse)
meuse.pbN<-meuse[seq(1, length(meuse$lead), by=3), c('x', 'y', 'lead', 'zinc')]
NfoldDiff<-vector()
for(i in 1:length(meuse.pb$lead)){
  #remove one case of the target variable
  meuse.pbn<-meuse.pbN[-i, c('x', 'y', 'lead', 'zinc')]
  meuse.pbn<-cbind(meuse.pbn, ltpb=log10(meuse.pbn$lead), ltzn=log10(meuse.pbn$zinc))
  meuse.pb.extra <- meuse.pbN[i, c("x", "y", "lead")]
  meuse.pb.extra <- cbind(meuse.pb.extra, ltpb = log10(meuse.pb.extra$lead))
  coordinates(meuse.pbn) <- ~ x + y
  coordinates(meuse.pb.extra) <- ~ x + y
  #variogram construction on n-1 data points
  v.ltpb<-variogram(ltpb~1, data=meuse.pbn, cutoff=1800, width=200)
  m.ltpb<-vgm(psill=0.08, model='Sph', range =800, nugget=0.03)
  m.ltpb.f <- fit.variogram(v.ltpb, m.ltpb)
  #conduct ordinary kriging
  k <- krige(ltpb ~ 1, meuse.pbn, meuse.pb.extra, m.ltpb.f)
  NfoldDiff[i] <- meuse.pb.extra$ltpb - k$var1.pred  
}
summary(NfoldDiff)
MSEmyCVdiff<-sqrt(sum(NfoldDiff^2)/length(NfoldDiff))
BiasmyCVdiff<-sum(NfoldDiff)/length(NfoldDiff)     # mean error (bias)
medmyCVdiff<-median(NfoldDiff)
```



```{r CoKriging,message=FALSE,results='hide'}
#correlation
cor(meuse.pb$ltzn, meuse.pb$ltpb)
plot(meuse.pb$ltzn, meuse.pb$ltpb)
#extract the co-variate
meuse.co<-as.data.frame(meuse)[, c('x', 'y', 'zinc')]
meuse.co <- cbind(meuse.co, ltzn = log10(meuse.co$zinc))
coordinates(meuse.co) <- ~ x + y
#variogram
v.ltzn <- variogram(ltzn ~ 1, meuse.co, cutoff=1800)
plot(v.ltzn, pl=T)
m.ltzn <- vgm(psill=.11, model="Sph", range=1000, nugget=.02)  #by visual inspection
m.ltzn.f <- fit.variogram(v.ltzn, m.ltzn)         #by least square fit
plot(v.ltzn, pl=T, model=m.ltzn.f)
#Model the coregionalization: create a gstat structure containing sets of data: subsample for lead 
#and full sample for zinc
g1 <- gstat(NULL, id = "ltpb", form = ltpb ~ 1, data = meuse.pb)
g1 <- gstat(g1, id = "ltzn", form = ltzn ~ 1, data = meuse.co)
#variogram and cross-variogram
v.cross <- variogram(g1)
plot(v.cross, pl=T)
#Fit cross-variogram model
g1 <- gstat(g1, id = "ltpb", model = m.ltpb.f, fill.all=T)
g1 <- fit.lmc(v.cross[-16,], g1)
plot(variogram(g1), model=g1$model)

#Prediction at the grids
k.c1 <- predict.gstat(g1, meuse.grid)
summary(k.c1$ltpb.pred); summary(k.c1$ltpb.var)
#Add to DF for plotting
meuse.grid$k.cl<-k.c1$ltpb.pred
meuse.grid$k.cl.se<-sqrt(k.c1$ltpb.var)
##Validation
#Interpolate at extra points
k <- predict.gstat(g1, meuse.extra)
codiff <- k$ltpb.pred - meuse.extra$ltpb
summary(codiff)
MSEcodiff<-sqrt(sum(codiff^2)/length(codiff))     # RMS error (precision)
Biascodiff<-sum(codiff)/length(codiff)     # mean error (bias)
medcodiff<-median(codiff)         # median error

#Cross Validation
cv.c2 <- gstat.cv(g1, nfold=52)
pe2<-cv.c2$residual
# summary(cv.c2$residual)
MSEcvcodiff<-sqrt(mean(cv.c2$residual^2))
# mean(cv.c2$residual)
Biascvcodiff<-sum(cv.c2$residual)/length(cv.c2$residual)
medcvcodiff<-median(cv.c2$residual)
```

```{r CoKrigingWithom,message=FALSE,results='hide'}
#correlation
cor(meuse.pb$ltzn, meuse.pb$om,use="complete.obs")
plot(meuse.pb$ltzn, meuse.pb$om)
#extract the co-variate
meuse.co<-na.omit(as.data.frame(meuse)[, c('x', 'y', 'om')])
coordinates(meuse.co) <- ~ x + y
#variogram
v.om <- variogram(om ~ 1, meuse.co, cutoff=1800)
plot(v.om, pl=T)
m.om <- vgm(psill=.12, model="Sph", range=800, nugget=.02)  #by visual inspection
m.om.f <- fit.variogram(v.om, m.om)         #by least square fit
plot(v.om, pl=T, model=m.om.f)
#Model the coregionalization: create a gstat structure containing sets of data: subsample for lead 
#and full sample for zinc
g1 <- gstat(NULL, id = "ltpb", form = ltpb ~ 1, data = meuse.pb)
g1 <- gstat(g1, id = "om", form = om ~ 1, data = meuse.co)
#variogram and cross-variogram
v.cross <- variogram(g1)
plot(v.cross, pl=T)
#Fit cross-variogram model
g1 <- gstat(g1, id = "ltpb", model = m.ltpb.f, fill.all=T)
g1 <- fit.lmc(v.cross[-16,], g1)
plot(variogram(g1), model=g1$model)

#Prediction at the grids
k.c2 <- predict.gstat(g1, meuse.grid)
summary(k.c2$ltpb.pred); summary(k.c2$ltpb.var)
#Add to DF for plotting
meuse.grid$k.cl<-k.c1$ltpb.pred
meuse.grid$k.cl.se<-sqrt(k.c1$ltpb.var)
##Validation
#Interpolate at extra points
k2 <- predict.gstat(g1, meuse.extra)
codiff <- k2$ltpb.pred - meuse.extra$ltpb
summary(codiff)
MSEco2<-sqrt(sum(codiff^2)/length(codiff))     # RMS error (precision)
Biasco2<-sum(codiff)/length(codiff)     # mean error (bias)
medco2<-median(codiff)         # median error

#Cross Validation
cv.c2 <- gstat.cv(g1, nfold=52)
pe2.2<-cv.c2$residual
# summary(cv.c2$residual)
MSEcvco2<-sqrt(mean(cv.c2$residual^2))
# mean(cv.c2$residual)
Biascvco2<-sum(cv.c2$residual)/length(cv.c2$residual)
medcvco2<-median(cv.c2$residual)
```

#Assignment 1, 2 and 3 

For the ease of interpretation and comparison root-mean-squared errors and biases for all methods are presented in Table 1.  The estimate of prediction accuracy using the extra points is lower than the estimate usinf n-fold cross-validation.  I used the built in cross-validation method and manually coded cross-validation for the ordinary kriging prediction.  Results from these two methods were very similar but not identical.  The RMSE from n-fold cross validation is likely to be a better estimate of prediction error on a new a sample since it more systematically addresses the uncertainty.

The use of co-kriging with zinc greatly reduced the prediction error when compared to ordinary kriging.  The predictions and standard errors for ordinary kriging and co-kriging are given in figures 1 and 2.  These figures show the better prediction accuracy of co-kriging with zinc.  However, co-kriging with 'om' rather than zinc reduced the prediction accuracy (Table 1).  This is because the correlation between 'om' and lead was not nearly as strong as the correlation between zinc and lead (0.72 vs 0.97).  

```{r TableOfStats,results='asis'}
library(pander)
sumtab<-matrix(c(MSE_ok,MSE_okcv,MSEmyCVdiff,MSEcvcodiff,MSEcvco2,Bias_ok,Bias_okcv,BiasmyCVdiff,Biascvcodiff,Biascvco2),5,2)
colnames(sumtab)<-c("RMSE","Bias")
rownames(sumtab)<-c("Ordinary Kriging","Ordinary Kriging (CV)","Ord. Krig. (manual CV)","Co-kriging with Zinc","Co-kriging with 'om'")
pandoc.table(sumtab,caption="Root-mean-squared errors and biases for all methods.")
```



```{r SetUpPlots,echo=FALSE}
library(maptools) 
bluepal<-colorRampPalette(c('azure1','steelblue4')) 
brks<-c(0, 1.6, 1.8, 2, 2.2, 2.4, 2.6) 
cols<-bluepal(length(brks)-1) 
sepal<-colorRampPalette(c('peachpuff1', 'tomato3')) 
brks.se<-c(0.18,0.2,.22,.24,.26,.28,.3,.32) 
cols.se<-sepal(length(brks.se)-1) 
scols<-c('green', 'red') 
```

```{r Plotkclko,message=FALSE, fig.width=7.5,fig.height=7}
meuse.grid.df<-SpatialPixelsDataFrame(points=meuse.grid@coords,data=as.data.frame(meuse.grid))

#predict over entire area
OK_fit<-gstat(id='OK_fit', formula=ltpb ~ 1, data=meuse.pb, model=m.ltpb.f) 
pe<-gstat.cv(OK_fit, nfold=52, debug.level=0, random=FALSE)$residual 
ok<-predict(OK_fit, newdata=meuse.grid.df, debug.level=0) 
meuse.grid.df$ok_pred<-ok$OK_fit.pred
meuse.grid.df$ok_se<-sqrt(ok$OK_fit.var)

par(mfrow=c(1,2))
image(meuse.grid.df, 'ok_pred', breaks=brks, col=cols)
symbols(meuse.pbN[c("x","y")], circles=abs(pe)*100, fg='black', bg=scols[(pe<0)+1], inches=FALSE, add=TRUE)
legend('topleft',fill=cols, legend=leglabs(brks), bty ='n', cex=0.8)
legend('bottomright', fill=scols, legend=c("Negative Residual","Positive Residual"), bty='n', cex=0.8)
title("Ordinary Kriging","Predictions")

image(meuse.grid.df, 'k.cl', breaks=brks, col=cols)
symbols(meuse.pbN[c("x","y")], circles=abs(pe2)*100, fg='black', bg=scols[(pe2<0)+1], inches=FALSE, add=TRUE)
legend('topleft',fill=cols, legend=leglabs(brks), bty ='n', cex=0.8)
legend('bottomright', fill=scols, legend=c("Negative Residual","Positive Residual"), bty='n', cex=0.8)
title("Co-kriging","Predictions")
#plot SE's
image(meuse.grid.df, 'ok_se', breaks=brks.se, col=cols.se)
symbols(meuse.pbN[c("x","y")], circles=abs(pe)*100, fg='black', bg=scols[(pe<0)+1], inches=FALSE, add=TRUE)
legend('topleft',fill=cols.se, legend=leglabs(brks.se), bty ='n', cex=0.8)
legend('bottomright', fill=scols, legend=c("Negative Residual","Positive Residual"), bty='n', cex=0.8)
title("Ordinary Kriging","Standard Errors")

image(meuse.grid.df, 'k.cl.se', breaks=brks.se, col=cols.se)
symbols(meuse.pbN[c("x","y")], circles=abs(pe2)*100, fg='black', bg=scols[(pe2<0)+1], inches=FALSE, add=TRUE)
legend('topleft',fill=cols.se, legend=leglabs(brks.se), bty ='n', cex=0.8)
legend('bottomright', fill=scols, legend=c("Negative Residual","Positive Residual"), bty='n', cex=0.8)
title("Co-kriging","Standard Errors")
par(mfrow=c(1,1))

```